\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-04-14 archive.tex
deep program learning experiment analysis

\section{Deep Wake/Sleep Program Induction}\label{overviewSection}

Our model, called \systemEnding, works through a novel kind
of wake/sleep or `dream' learning~\cite{hinton1995wake}, iterating
through a wake cycle -- where it solves \textbf{tasks} by writing programs (Figure~\ref{threeCycles} top)
-- and a pair of sleep cycles, both of which are loosely biologically
inspired. The first sleep cycle, which we refer to as
\textbf{abstraction}, grows its library of code by replying
experiences from waking and consolidating them into new code
abstractions (Figure~\ref{threeCycles} left).  This cycle is inspired by the formation of abstractions
during sleep memory consolidation~\cite{DUDAI201520}.  The second
sleep cycle, which we refer to as \textbf{dreaming}, improves the
agents knowledge of how to write code by training a neural network to
help quickly search for programs. The neural net is trained on replayed
experiences as well as `fantasies', or samples, from the learned
library (Figure~\ref{threeCycles} right).  These two kinds of dreams are inspired by the distinct
episodic replay and hallucination components of dream
sleep~\cite{fosse2003dreaming}.

@@

The library offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions across the
programs found during waking.  Second, we use this automatic refactoring
process to search for libraries that maximally compress these programs by
incorporating reused subexpressions into the DSL.

@@

\subsubsection{List Processing}\label{listSection}
We took 218 list manipulation tasks from our previous work~\cite{ecc},
each with 15 input/output examples.  In solving these tasks, the
system typically composes around 20 new library routines, and discovered multiple
higher-order functions. Each round of memory consolidation built on
components discovered in earlier sleep cycles --- for example the
agent first learns the higher-order function \code{filter}, uses
\code{filter} to learn to take the maximum element of a list, then
uses that routine to learn a new component for extracting the
$n^{\text{th}}$ largest element of a list, which it finally uses to
solve a task involving sorting a list of numbers (Figure~\ref{initialExampleDSL}).
This incremental, modular learning of deep hierarchies of library
components occurs because of the alternation between code writing
(during waking) and code refactoring (during the abstraction phase
of sleep).

@@

Procedural visual concepts are studied across AI and cognitive science
--- Bongard problems~\cite{Moscow}, Raven's progressive
matrices~\cite{raven2003raven}, and Lake et al.'s BPL model of
omniglot~\cite{lake2015human} are prominent examples.  Here we take
inspiration from LOGO Turtle graphics~\cite{turtle}, tasking our agent
with drawing a corpus of 160 images while equipping it with control
over a `pen', along with arithmetic operations on angles and
distances.

@@

Inspired by the classic AI `copy task' --- where an agent must look at
an image of a tower made of toy blocks and re-create the
tower~\cite{towerCopy} --- we give \system 112 tower `copy tasks'
(Figure~\ref{towerTasks}).  Here the agent observes both an image of a
tower and the locations of each of its blocks, and must write a
program that plans how a simulated hand would build the tower.
These towers are built from Lego-style blocks
that snap together on a discrete grid.
The system starts with the same control flow primitives as with
LOGO graphics,
and learns parametric `options' for building blocks towers (Figure~\ref{tower}),
inferring concepts like arches, staircases, bridges and brick walls.
\begin{figure}
  \begin{tabular}{cc}
    \toprule Example learned DSL components&Samples from DSL\\\midrule
    \begin{tabular}{rl}
      Brickwall&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bricks.png}
      \end{tabular}\\
      Bridge&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bridge.png}
      \end{tabular}\\
      Staircase&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_staircase.png}
        \end{tabular}
    \end{tabular}&
    \begin{tabular}{l}
      \includegraphics[width = 5cm]{figures/tower/dreams.png}
      \end{tabular}
    \\\bottomrule 
    \end{tabular}
  \caption{\textbf{Left}: Three (out of 19) learned DSL components for building towers out of Lego-style blocks. These components act like parametric options~\cite{stolle2002learning},
    giving higher-level building blocks that the agent can use to plan. \textbf{Right}: 16 random samples, or `dreams', built from learned library.}\label{tower}
\end{figure}

@@

\subsection{Quantitative Results on Held-Out Tasks}\label{quantitative}
To evaluate the relative importance of library learning and recognition
model training, we evaluate on held-out testing tasks for each of our
domains, measuring both how many tasks are solved and how long it
takes to solve them across successive wake/sleep iterations
(Fig.~\ref{learningCurves}).  We always solve more held-out tasks --
and generally solve them in less time -- with both components
combined.  Why?  One hypothesis is that some tasks are best solved by
DSL learning and others by a neural network, and so including both of
our sleep cycles  takes the union of these sets of tasks.  Another
hypothesis is that the DSL and neural network interact
synergistically, bootstrapping off each other to solve tasks for which
neither alone suffice. We evaluate the relative weight of these interactions by
computing the ratio of the tasks solved uniquely by the full model to

\begin{figure}
  \begin{tabular}{ccc}
    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/text_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/text_time.png}
    \end{tabular}&

@@

@@

\emph{(what we found)} Our goal was to build a model capturing
humanlike aspects of learning to be an expert in a new domain of
problems.  Two main findings of our work are that, first, a single
program-induction system can learn to solve large sets of problems
from many qualitatively different domains; and second, that fully
acquiring expertise in these domains hinges both upon explicit
declarative knowledge, and implicit procedural skill.  We have
deliberately placed the interplay of declarative and procedural
knowledge at the center of our model, and believe that both these
kinds of learning are crucial for building agents that, like humans,
autonomously learn to navigate a new domain of problems.

@@

\subsection{Full set of LOGO  tasks}

\begin{figure}
  \includegraphics[width = \textwidth]{figures/fullLogo.png}
  \caption{Full set of LOGO graphics tasks that we apply our system to}
\end{figure}

\subsection{Full set of tower tasks}
\begin{figure}
  \includegraphics[width = \textwidth]{figures/fullTower.png}
  \caption{Full set of tower building tasks that we apply our system to}
\end{figure}

>>>>

\section{A Wake/Sleep Algorithm for Program Induction}\label{overviewSection}

Our model, called \systemEnding, works through a novel kind
of wake/sleep or `dream' learning~\cite{hinton1995wake}, iterating
through a wake cycle -- where it solves \textbf{tasks} by writing programs (Figure~\ref{threeCycles} top)
-- and a pair of sleep cycles, both of which are loosely biologically
inspired by the different memory consolidation processes that occur during sleep~\cite{rasch2013sleep}. The first sleep cycle, which we refer to as
\textbf{abstraction}, grows the library of code by replying
experiences from waking and consolidating them into new code
abstractions (Figure~\ref{threeCycles} left).  This cycle is loosely inspired by the formation of declarative abstractions
during slow-wave sleep memory consolidation~\cite{DUDAI201520}.  The second
sleep cycle, which we refer to as \textbf{dreaming}, improves the
agents procedural knowledge of how to write code by training a neural network to
help quickly search for programs. The neural net is trained on replayed
experiences as well as `fantasies', or sampled programs, built from the learned
library (Figure~\ref{threeCycles} right).  These two kinds of dreams are inspired by the distinct
episodic replay and hallucination components of dream
sleep~\cite{fosse2003dreaming}.

@@

The library offers a set of abstractions that allow an agent to concisely
express solutions to the tasks at hand. We automatically discover
these new abstractions by combining two ideas. First, we build on
techniques from the programming languages community to develop a new
algorithm for automatically refractoring programs, where this
refactoring exposes common reused subexpressions across the
programs found during waking.  Second, we use this automatic refactoring
process to search for libraries that maximally compress these programs by
incorporating reused subexpressions into the library.

@@

\subsubsection{List Processing}\label{listSection}
We took 218 list manipulation tasks from our previous work~\cite{ecc},
each with 15 input/output examples.  In solving these tasks, the
system typically composes around 20 new library routines, and discovered multiple
higher-order functions. Each round of memory consolidation built on
components discovered in earlier sleep cycles --- for example the
agent first learns the higher-order function \code{filter}, uses
\code{filter} to learn to take the maximum element of a list, then
uses that routine to learn a new component for extracting the
$n^{\text{th}}$ largest element of a list, which it finally uses to
solve a task involving sorting a list of numbers (Figure~\ref{exampleDSL}).
This incremental, modular learning of deep hierarchies of library
components occurs because of the alternation between code writing
(during waking) and code refactoring (during the abstraction phase
of sleep).

@@

Procedural or generative visual concepts 
--- from Bongard problems~\cite{Moscow}, to handwritten characters~\cite{lake2015human,hofstadter1993letter}, to Raven's progressive
matrices~\cite{raven2003raven} --- are studied across AI and cognitive science.  Here we take
inspiration from LOGO Turtle graphics~\cite{turtle}, tasking our agent
with drawing a corpus of 160 images while equipping it with control
over a `pen', along with arithmetic operations on angles and
distances.

@@

Inspired by the classic AI `copy task' --- where an agent must look at
an image of a tower made of toy blocks and re-create the
tower~\cite{towerCopy} --- we give \system 112 tower `copy tasks'
(Figure~\ref{towerTasks}).  Here the agent observes both an image of a
tower and the locations of each of its blocks, and must write a
program that plans how a simulated hand would build the tower.
These towers are built from Lego-style blocks
that snap together on a discrete grid.
The system starts with the same control flow primitives as with
LOGO graphics,
and learns parametric `options' for building blocks towers (Figure~\ref{tower}),
inferring concepts like arches, staircases, bridges and brick walls.
\begin{figure}
\centering  \begin{tabular}{cc}
    \toprule Example learned DSL components&Samples written using final library\\\midrule
    \begin{tabular}{rl}
      Brickwall&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bricks.png}
      \end{tabular}\\
      Bridge&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_bridge.png}
      \end{tabular}\\
      Staircase&\begin{tabular}{l}
        \includegraphics[width = 5cm]{figures/tower/tower_dsl_staircase.png}
        \end{tabular}
    \end{tabular}&
    \begin{tabular}{l}
      \includegraphics[width = 5cm]{figures/tower/dreams.png}
      \end{tabular}
    \\\bottomrule 
    \end{tabular}
  \caption{\textbf{Left}: Three (out of 19) learned DSL components for building towers out of Lego-style blocks. These components act like parametric options~\cite{stolle2002learning},
    giving higher-level building blocks that the agent can use to plan. \textbf{Right}: 16 random samples, or `dreams', built from learned library.}\label{tower}
\end{figure}

@@

\subsection{Quantitative Results on Held-Out Tasks}\label{quantitative}
To evaluate the relative importance of library learning and recognition
model training, we evaluate on held-out testing tasks for each of our
domains, measuring both how many tasks are solved and how long it
takes to solve them across successive wake/sleep iterations
(Fig.~\ref{learningCurves}).  We always solve more held-out tasks --
and generally solve them in less time -- with both components
combined, suggesting a synergy between library learning and
recognition model training. Examining the libraries learned with and without
recognition models reveals a difference in the typical \emph{depth}
of learned subroutines, and this average depth correlates
with the number of solved testing tasks ($r=0.71$, $p < 10^{-8}$).
In contrast, the correlation between the \emph{number} of learned subroutines and number of solved tasks is weaker and not statistically significant ($r=0.20$, $p > 0.05$),
suggesting that the recognition model bootstraps
``better'' libraries, where ``better'' is at least partly explained by the depth
of the learned representation. Figure~\ref{depthMontage} shows typical `deep' libraries learned by
\system and contrasts this depth with \% solved tasks, both with and without the recognition model.
\begin{figure}
  \begin{tabular}{ccc}
    \begin{tabular}{c}
      \includegraphics[width = 5cm]{figures/learningCurves/text_hits.png}\\
      \includegraphics[width = 5cm]{figures/learningCurves/text_time.png}
    \end{tabular}&

@@

\begin{figure}
  \includegraphics[width = \textwidth]{figures/dpl/montage.pdf}
  \caption{Top: Learned libraries diagrammed as networks. Arrows point from a learned concept to all other concepts that use it. Bottom: Deeper networks solve more testing tasks, both as measured by average depth (left) and maximum depth (right). Full model (circles on scatterplots) learns deeper libraries than no-recognition-model ablation (crosses on scatterplots). Each scatter plot point run with a different random seed.}\label{depthMontage}
  \end{figure}

@@

\emph{(what we found)} Our goal was to build a model capturing
humanlike aspects of learning to be an expert in a new domain of
problems. Our model works through a wake/sleep algorithm that grows an increasingly deep
representation of a domain's conceptual structure,
while learning to quickly combine these concepts into programs
solving specific problems within the domain.
Two main findings of our work are that, first, a single
program-induction system can learn to solve large sets of problems
from many qualitatively different domains; and second, that fully
acquiring expertise in these domains hinges both upon explicit
declarative knowledge, and implicit procedural skill.  We have
deliberately placed the interplay of declarative and procedural
knowledge at the center of our model, and believe that both these
kinds of learning are crucial for building agents that, like humans,
autonomously learn to navigate a new domain of problems.

@@

\subsection{LOGO Graphics Experiment}

We apply our system to 160 LOGO graphics tasks, split 50/50 test/train (Figure~\ref{everyLogoTask}).
For each task of the agent must write a program that moves a simulated pen across the canvas,
with the goal of drawing the target image, as measured by pixel level similarity.

We initially provide our agent with two control flow primitives: \code{for} (a `for` loop), and \code{get/set}, which gets and saves the current state of the agent's pen, executes a block of code, and then restores the state to its previous value. We provide two different drawing primitives: \code{move}, which takes as input both a distance and angle, and moves the pen forward by that distance and rotates by that angle, and \code{pen-up}, which lifts up the pen and then executes a block of code \emph{without} placing down ink on the canvas.

These programs are imperative (produce a sequence of actions) rather than purely functional (calculate a stateless mapping).
The \system infrastructure only supports purely functional $\lambda$-calculus programs.
To embed imperative routines within this framework,
we model imperative routines using a state monad~\cite{wadler1990comprehending} and encode each imperative action in a continuation passing style where each imperative primitive takes as input the actions to execute afterwards (i.e., the continuation),
and the program as a whole takes as input a final no-op continuation.
This continuation passing set up allows the program synthesizer to not need the monadic `bind' operator to sequence chains of commands.
To be concrete, our imperative primitives have the following types:

\begin{tabular}{l}
  \code{move} : \code{distance}$\to$\code{angle}$\to$\code{State ()}$\to$\code{State ()}\\
  \code{pen-up} : \code{(State ()$\to$State ())}$\to$\code{State ()}$\to$\code{State ()}\\
  \code{for} : \code{int$\to$(State ()$\to$State ())}$\to$\code{State ()}$\to$\code{State ()}\\
  \code{get/set} : \code{(State ()$\to$State ())}$\to$\code{State ()}$\to$\code{State ()}
\end{tabular}

and each synthesized program has type \code{State ()}$\to$\code{State ()}. We additionally provide the agent with natural numbers (1 through 9), arithmetic operations (addition/subtraction/multiplication/division), unit distances and angles (1 meter and $2\pi$ radians), and special values $\infty$ (approximated as $20$) and $\epsilon$ (approximated as $1/20$). We distinguish distances and angles in the type system, only allowing a distance/angle to be added/subtracted with another object that is also a distance/angle, and only allowing multiplication/division by unitless scalars. This trick requires no modification of the underlying \system software, and is a standard way of embedding units and dimensions within a type system~\cite{fsunits}.

\begin{figure}
  \includegraphics[width = \textwidth]{figures/fullLogo.png}
  \caption{Full set of LOGO graphics tasks that we apply our system to}\label{everyLogoTask}
\end{figure}

\subsection{Tower Building Experiment}
We apply our system to 112 tower building tasks (Figure~\ref{everyTowerTask}), split 50/50 test/train.
We use the same monadic, continuation-passing encoding of imperative programs as we used for LOGO graphics,
and include the exact same control flow primitives.
Rather than moving a pen over a canvas, the agent here moves a simulated hand left/right over a 2D tower building stage, and has at its disposal an unlimited supply of horizontal and vertical blocks.
All coordinates are discretized.
Horizontal blocks have size 6x2 and vertical blocks have size 2x6.
The state of the simulated hand maintains both its position (a single discrete scalar) and its orientation (a single boolean, either facing left are facing right).
We include two domain specific primitives for adjusting the position of the hand: \code{move}, which takes as input a scalar $d$ and moves the hand forward  by distance $d$,
and \code{reverse},
which flips the orientation of the hand.
The agent has two additional domain specific primitives for dropping a horizontal/vertical block at the current position of the hand.

\begin{figure}
  \includegraphics[width = \textwidth]{figures/fullTower.png}
  \caption{Full set of tower building tasks that we apply our system to}\label{everyTowerTask}
\end{figure}
